Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Updatse

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative DAG construction that
combines domain knowledge with data-driven insights. Our approach iteratively
identifies pair of variables with a missing edge in the DAG, and asks for the
orientation of the edge from an expert. We also propose a method to rank these
potential new edges based on their impact on the model fit. Empirical results
show that this iterative model construction approach achieves performance
comparable to automated algorithms if we are able to correctly decide the
orientation of new edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model (LLM) can
be used to determine edge orientations. We provide the implementation of our
approach in both a web tool at: <redacted for review> and a Python package
<redacted for review>.

@@

Figure~\ref{fig:intro} highlights some of these issues.

When manually constructing models, it is important to test whether the
resulting DAG accurately represents the dataset. Based on this evaluation and
domain knowledge, we can make further modification to the model. One common
approach for assessing DAGs is to check whether the Conditional Independences
(CIs) implied by the DAG hold in the data \citep{Ankan2021}. If we find
violations to these CIs, we can use our domain knowledge to make appropriate
modifications to the DAG. However, this CI testing approach has a few
limitations:

\begin{enumerate}
\item As CIs are only implied by missing edges in the model, this approach does 
	not test whether the existing edges are correct.
\item Determining whether a CI holds in data is based on a p-value threshold. This can 
	be unreliable, for example, we get a significant p-value for even very weak 
	relationship if the sample size is high.
\item The number of implied CIs can be large making it difficult to manually
	check all of them. For example, a moderately sized alarm network
	\citep{Beinlich1989} with $ 37 $ nodes, implies $ 287 $ CIs.
\end{enumerate}

To tackle these limitations, we draw inspiration from modification indices in
specification search. Modification indices provide ranking for potential model
modification based on their impact on the model's fit, allowing us to
prioritize most meaningful modifications.  Similarly, we propose a method for
ranking CI test violations to help us focus on the most critical
inconsistencies in the model. To determine this ranking we utilize a measure of
conditional association between variables in combination with p-values from the
CI test. This approach addresses all three of the issues outlined above. The
measure of association allows us to evaluate the validity of existing edges. It
reduces our reliance on p-values alone for deciding whether the CI test holds.
The ranking helps us prioritize the most critical violations, making it feasible 
to focus on the areas that most improve the model's fit.

@@

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative DAG construction method that
combines domain knowledge with data-driven insights. Our approach iteratively
identifies pair of variables with a missing edge in the DAG, and asks for the
orientation of the edge from an expert. We also propose a method to rank these
potential new edges based on their impact on the model fit. Empirical results
show that this iterative model construction approach achieves performance
comparable to automated algorithms if we are able to correctly decide the
orientation of new edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model (LLM) can
be used to determine edge orientations. We provide the implementation of our
approach in both a web tool at: <redacted for review> and a Python package
<redacted for review>.

@@

Figure~\ref{fig:intro} highlights some of these issues. To assist researchers
in constructing DAGs manually, we propose an iterative model construction
method that combines domain knowledge with data-driven insights. Our approach 
iteratively identifies pairs of variables with a missing edge in the DAG and

% When manually constructing models, it is important to test whether the
% resulting DAG accurately represents the dataset. Based on this evaluation and
% domain knowledge, we can make further modification to the model. One common
% approach for assessing DAGs is to check whether the Conditional Independences
% (CIs) implied by the DAG hold in the data \citep{Ankan2021}. If we find
% violations to these CIs, we can use our domain knowledge to make appropriate
% modifications to the DAG. However, this CI testing approach has a few
% limitations:
% 
% \begin{enumerate}
% \item As CIs are only implied by missing edges in the model, this approach does 
% 	not test whether the existing edges are correct.
% \item Determining whether a CI holds in data is based on a p-value threshold. This can 
% 	be unreliable, for example, we get a significant p-value for even very weak 
% 	relationship if the sample size is high.
% \item The number of implied CIs can be large making it difficult to manually
% 	check all of them. For example, a moderately sized alarm network
% 	\citep{Beinlich1989} with $ 37 $ nodes, implies $ 287 $ CIs.
% \end{enumerate}
% 
% To tackle these limitations, we draw inspiration from modification indices in
% specification search. Modification indices provide ranking for potential model
% modification based on their impact on the model's fit, allowing us to
% prioritize most meaningful modifications.  Similarly, we propose a method for
% ranking CI test violations to help us focus on the most critical
% inconsistencies in the model. To determine this ranking we utilize a measure of
% conditional association between variables in combination with p-values from the
% CI test. This approach addresses all three of the issues outlined above. The
% measure of association allows us to evaluate the validity of existing edges. It
% reduces our reliance on p-values alone for deciding whether the CI test holds.
% The ranking helps us prioritize the most critical violations, making it feasible 
% to focus on the areas that most improve the model's fit.

@@

\subsection{Integrating domain knowledge in causal discovery}
